---
title: "Quantitative methods - solutions to exercises"
author: "Katrín Björnsdóttir & Camila Pacheco Riaño"
date: "2024-04-26"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1. Hierarchical linear model exercise

## Libraries

*Remeber that you have to load the packages each time you start a new R session*

```{r, warning = FALSE, meassage = FALSE}
library(lme4)
library(tidyverse)
library(ggeffects) 
```

## Load the data

```{r, message = FALSE}

itex <- read_csv("Data/ITEX_diversity_data.csv") %>% 
  dplyr::select(-...1) # this code takes out the first column that was called "...1", it's not necessary though, it just makes the dataframe cleaner I think

```

Lets look at the data to make sure it looks okay

```{r}
head(itex)
```

## Exercise

What is the relationship between diversity and temperature across sites (i.e. across latitude)?

```{r}
itex <- itex %>% 
  mutate(TempC = (WarmQuarterTemp - 30)/2) #change temperature from F to °C (optional)
```

mutate is used to make a new column, in this case we are making a new column called "TempC" by subtracting 30 from the column "PlotTemp" and dividing by 2.

```{r}
hist <- hist(itex$SppRich) # this shows us the distribution of our response variable
```

It is a good practice to scale the explanatory variables.

```{r}
itex <- itex %>% 
  mutate(TempC_scaled = scale(TempC, center = TRUE, scale = TRUE))
```

Do you remember what that does?

.....

By scaling/centering you position the data around the mean of 0. So the values will be positioned on either side of 0 depending on how far from the mean the values are.

By exploring the relationship between diversity and temperature across sites we could start with a simple linear model.

```{r}
itex.lm <- lm(SppRich ~ TempC_scaled, data = itex)
summary(itex.lm)
```

From the summary output you can read plenty, but I'll explain a few for now. -

-   The estimate is the slope of the relationship (1.7669)

-   Pr is the p value from the model - in our case it is significant since it is lower than 0.05.

-   Adjusted R-squared tells us how much variation was explained by the model - in our case 0.1067 or 10%.

Now lets plot this relationship

```{r}
(prelim_plot <- ggplot(itex, aes(x = TempC_scaled, y = SppRich)) +
    geom_point() +
    geom_smooth(method = "lm"))
```

Notice that we get a warming message ... wonder what is causing that?

To evaluate how good the model fit was we can look at the residuals (the variation that was not explained by the model) against the fitted values. This is done to detect non-linearity, unequal error variances and outliers. When a linear regression model is suitable for a data set, then the residuals are more or less randomly distributed around the 0 line.

```{r}
plot(itex.lm, which = 1)  # not perfect... but we'll go with it
```

Now lets explore this relationship a bit more, lets add colors based on sites to see if we notice any pattern.

*sidenote : in ggplot you can add colors by including "color = SITE" within the aesthetic*

```{r, warnings = FALSE}
(colour_plot <- ggplot(itex, aes(x = TempC_scaled, y = SppRich, color = SITE)) +
    geom_point(size = 2) +
    theme_classic() +
    theme(legend.position = "none"))
```

Can we see any pattern? ... Yes I think so. The colors which represent different sites are in some cases clustered together.

Because of the nested structure of our data, i.e. plots within subsites within sites we might want to think about using a **linear mixed model**. Then we need to think about what we want to include as random effects. Because we expect plots within SITE and SUBSITE to be more similar to each other, we will add SITE and SUBSITE as random effect

```{r}
mixed.lmer <- lmer(SppRich ~ TempC_scaled + (1|SITE/SUBSITE), data = itex)
summary(mixed.lmer)
```

Lets first look at if the summary output corresponds to the structure of the data.

From the output we can see that the model was fitted with: 1294 obs., 82 SUBSITES and 23 SITES

*see this line in the summary output - Number of obs: 1294, groups: SUBSITE:SITE, 82; SITE, 23*

Hmmmm... that's not quite what we have is it?

```{r}
nrow(itex) # we have 1303 observation
```

```{r}
length(unique(itex$SITE)) # 24 sites
```

```{r}
length(unique(itex$SUBSITE)) # 83 subsites
```

Something is not quite right here. Do you remember how we got a warning message earlier in the script when we were plotting the relationship?

The problem we have is that one of the sites has missing values that we should remove before the analysis

Lets look for missing values then

```{r}
sum(is.na(itex$SppRich)) # no NAs in SppRich, how about TempC_scaled?
```

```{r}
sum(is.na(itex$TempC_scaled)) # yes, here we have 9 missing values
```

```{r, warning = FALSE}
itex <- itex %>% 
  filter(!is.na(TempC_scaled)) # here we filter out all NA values within TempC_scaled
```

*sidenote: adding a "!" infront will remove values when using the filter function*

Alright lets try it again.

```{r}
nrow(itex) # now we have 1294 observation
```

```{r}
length(unique(itex$SITE)) # 23 sites
```

```{r}
length(unique(itex$SUBSITE)) # 82 subsites
```

Great now we can continue.

Run the model again with the fixed dataset.

```{r}
mixed.lmer <- lmer(SppRich ~ TempC_scaled + (1|SITE/SUBSITE), data = itex)
summary(mixed.lmer)
```

The summary output for linear mixed models in the lme4 package don't show p-values.This is a conscious choice made by the authors of the package, as there are many problems with p-values. Please be careful when it comes to evaluating the significance of your model. Focus on your question and avoid plugging in and dropping variables from a model haphazardly until you make something “significant”. Always choose variables based on biology/ecology and avoid adding in all possible variables that you have access to (i.e. don’t overfit). Remember that as a rule of thumb, you need 10 times more data than parameters you are trying to estimate. There are some common different methods used for model selection which we won't go into here but take a look at this thread if you are interested:
https://stats.stackexchange.com/questions/95054/how-to-get-an-overall-p-value-and-effect-size-for-a-categorical-factor-in-a-mi 

For simplification, we'll use the stargazer package that gives us a summary output table from our model.

```{r, message = FALSE}
#remember to install it first if you haven't already - then load!
library(stargazer)
```

```{r, warning = FALSE}
stargazer(mixed.lmer,
          digits = 3,
          type="text",
          star.cutoffs = c(0.05, 0.01, 0.001),
          digit.separator = "")
```

From this table we see that there is not a significant relationship with temperature and species richness across sites. If there would have been a significant relationship, there would have been an asterick/s after the 1.714 value (the slope of relationship).

Now for fun, lets visualize the new model to compare with our original linear model

Extract the prediction dataframe.

```{r}
pred.mm <- ggpredict(mixed.lmer, terms = c("TempC_scaled"))  # this gives an overall predictions for the model.

```

Now lets try to plot the predictions.

```{r}
#install.packages("viridis") #install this package to be able to add nice colors
```

*Sidenote: Only install this if you haven't done that previously, otherwise go ahead and load it.*

```{r, warning = FALSE}
library(viridis)
```

```{r}
(ggplot(pred.mm) + 
    geom_point(data = itex, 
               aes(x = TempC_scaled, y = SppRich, colour = SITE)) + # adding the raw data (scaled values) as points
    geom_line(aes(x = x, y = predicted), linewidth = 1, color = "black") +  # this is the slope
    geom_ribbon(aes(x = x, ymin = predicted - std.error, ymax = predicted + std.error), 
                fill = "lightgrey", alpha = 0.5) +  # this is the error band
    labs(x = "Temperature (°C)", y = "Species richness", 
         title = "") + 
    scale_color_viridis(discrete = TRUE, option = "viridis") + #this will add nice colors to the plot from the viridis package
    theme_classic() +
    theme(legend.position = "none") # here we are removing the legend
)

```

Now we can compare the two models to see how they differ visually.

```{r}
summary(itex.lm) #this is the output from the linear model
```

```{r}
(colour_plot <- ggplot(itex) +
    geom_point(size = 2, aes(x = TempC_scaled, y = SppRich, color = SITE)) +
    geom_smooth(method = "lm", aes(x = TempC_scaled, y = SppRich), color = "black") +
    scale_color_viridis(discrete = TRUE, option = "viridis") +
    theme_classic() +
    theme(legend.position = "none"))
```

Can you make sense of the difference? Notice how the slopes are very similar (1.7669 for the linear and 1.714 for the linear mixed model). What is different is the variation across sites. By adding a random effect to our model, we no longer detected a positive relationship with species richness and temperature across sites because of the increased uncertainty.

Now this is it for linear mixed modelling - hope you've learned something :-)

Remember to check out the Coding Club's INTRODUCTION TO LINEAR MIXED MODELS, find it here: <https://ourcodingclub.github.io/tutorials/mixed-models/>

and this is also a great source to understand how to work with colors in ggplot: <https://r-graph-gallery.com/ggplot2-color.html>

# 2. Ordinations

## Libraries

```{r, warning = FALSE, message = FALSE}
library(vegan)
```

## Exercise

Using the ITEX dataset (itex.long.csv), plot the species composition from two sites and four sub-sites in an NMDS ordination space and group them based on their location.

Load the data

```{r}
itex.long <- read_csv("Data/itex.long.csv") %>% 
  dplyr::select(-...1)
```

Lets look at the data

```{r}
head(itex.long)
```

Here we have another ITEX dataset where we have information on species composition and species abundance within two different ITEX sites.

```{r}
unique(itex.long$SITE) # two sites
```

```{r}
unique(itex.long$SUBSITE) # four sub-sites or habitats, two within each site
```

This is what the different habitat names represent:

-   AD = ATQASUK dry

-   AW = ATQASUK wet

-   BD = BARROW dry

-   BW = BARROW wet

First we need to convert our data from a long format to a wide format.

```{r}
itex.wide <- itex.long %>% 
  pivot_wider(names_from  = SPECIES_NAME, values_from  = SumAbund, values_fill = 0) 
```

Lets take a look at the dataset to see how this turned out

```{r}
head(itex.wide)
```

It looks good, our species are now columns and the species abundance within each plot are now rows.

Now remember how ordinations only take in numerical values, so we need to remove all columns that are something but numerical.

Lets take a look at the structure of out dataset

```{r}
str(itex.wide)
```

We have 101 columns (or variables) but the three first ones are characters. Lets remove them.

```{r}
itex.nmds <- itex.wide %>% 
  dplyr::select(-SITE, -SUBSITE, -PLOT)

```

Did that work?

```{r}
str(itex.nmds)
```

Yes that seemed to work

*Sidenote: Make sure that you make a new dataframe (in our case itex.nmds) instead of overwiting the old one, because we need information on the sites and subsites later on.*

Now we can start making our NMDS

First we assigning a distance matrix to "dist" using Bray-Curtis index (remember how that index was suitable for species abundance data).

```{r}
dist <- vegdist(itex.nmds,  method = "bray")
```

*Sidenote: if you get an error saying "Error in vegdist: (itex.nmds, method = "bray") : could not find function "vegdist" -- you need to load the vegan package !!*

Functions are used in R to perform a certain tasks. First you define the function (in our case NMDS.scree) and then later on you can call that function to perform that specific task.

```{r}
NMDS.scree <- function(x) { # where x is the name of the data frame variable
  plot(rep(1, 10), replicate(10, metaMDS(x, autotransform = F, k = 1)$stress), xlim = c(1, 10),ylim = c(0, 0.30), xlab = "# of Dimensions", ylab = "Stress", main = "NMDS stress plot")
  for (i in 1:10) {
    points(rep(i + 1,10),replicate(10, metaMDS(x, autotransform = F, k = i + 1)$stress))
  }
}
```

Here we are making a function to perform a NMDS for 1-10 dimensions and then plotting the nr. of dimensions vs the stress

We do this to evaluate the optimal number of dimensions we will use in our analysis.

```{r, message = FALSE, warning = FALSE, results = 'hide'}
NMDS.scree(dist)
```

This could take a while...

Now remember how we want the stress to be as low as possible but we also want as low number of dimensions as possible. From the plot we made we can see that there really isn't much difference between 2 and 3 - so we can go ahead and perform an NMDS using 2 dimensions.

set.seed() allows you to control the randomness in your code, making your results reproducible. It doesn't really matter what number you put in the brackets, you just have to put some number.

```{r}
set.seed(2)
```

Here, we perform the final analysis and check the result.

```{r}
NMDS1 <- metaMDS(dist, k = 2, trymax = 100, trace = F) 
```

k = 2 defines the number of dimensions we want, in our case 2, trymax determines the number of fits the ordination will perform to find the best fit, in our case 100 fits.

Lets check the results - check if the stress looks good.

```{r}
NMDS1
```

```{r}
stressplot(NMDS1)
```

Here 0.05 seems low enough.

Now we can plot our ordination.

```{r}
plot(NMDS1, type = "t")
plot(NMDS1,)
```

But notice how we only have our plots in this space, no species.

We can add the species scores by giving the metaMDS the original community matrix as input and specifying the distance measure. So essentially we are performing a new ordination analysis and instead of using vegsdist we define the distance within the ordination.

```{r}
NMDS2 <- metaMDS(itex.nmds, k = 2, trymax = 100, trace = F, autotransform = FALSE, distance = "bray")
plot(NMDS2)
plot(NMDS2, display = "sites", type = "n")
points(NMDS2, display = "sites", col = "red", cex = 1.25)
text(NMDS2, display ="species")
```

Alternatively, you can use the functions ordiplot and orditorp.

```{r}
ordiplot(NMDS2, type = "n")
orditorp(NMDS2, display = "species", col = "red", air = 0.01)
orditorp(NMDS2, display = "sites", cex = 1.1, air = 0.01)
```

Now this is a nice plot - we can see how the plots are really separated in space, each one of the four associated with different plant species.

But we don't know which group belongs to which sub-site or habitat.

First we need to define our subsites in a vector.

```{r, warning = FALSE}
group <- itex.wide$SUBSITE  #remember how I told you not to overwrite this dataset, now we are going to use the SUBSITE column to define our groups
colors <- c(rep("#ff0000", 24), rep("#590004", 24), rep("#606c38", 24), rep("#283618", 24)) #make a vector of different colors that you think are nice

```

Plot convex hulls with colors based on the group identity.

```{r, warning = FALSE}
ordiplot(NMDS2, type = "n")
#orditorp(NMDS2, display = "sites", type = "n", air = 0.01, cex = 1.25)
orditorp(NMDS2, display = "species", col = "#444444", air = 0.01)
for(i in unique(group)) {
  ordihull(NMDS2$point[grep(i, group),], draw="polygon",
           groups = group[group == i],col = colors[grep(i,group)],label=T) } 

```

You can add the plot numbers and the species names but you can also skip it because it makes the plot a bit complicated.. it's up to you :-)

Okay there we have it, notice how the habitats (the two dry habitats and the two wet habitats) are closer together in space than the two sites - why could that be?

Remember to check out the Coding Club's INTRODUCTION TO ORDINATION, find it here: <https://ourcodingclub.github.io/tutorials/ordination/>

Don't hesitate to contact us if you are having problems with running the script.

[katrin.bjornsdottir\@bioenv.gu.se](mailto:katrin.bjornsdottir@bioenv.gu.se){.email}

[camila.pacheco.riano\@bioenv.gu.se](mailto:camila.pacheco.riano@bioenv.gu.se){.email}

Good luck :-)
